Forecasting short-term hourly Emergency Departement arrivals

### 2021-11-05 paper/paper.Rmd
Started editing the paper

---
title:  Forecasting short-term hourly Emergency Departement arrivals
author:
  - name: Author1
    email: email1@example.com
    affiliation: University1
    footnote: 1
  - name: Jethro Browell
    email: jethro.browell@glasgow.ac.uk
    affiliation: School of Mathematics and Statistics, University of Glasgow, UK
    footnote: 2
  - name: Ivan Svetunkov
    email: i.svetunkov@lancaster.ac.uk
    affiliation: Centre for Marketing Analytics and Forecasting, Lancaster University, UK
    footnote: 2
address:
  - code: University1
    address: Cardiff business school, 3 Colum Drive, CF10 3EU, Cardiff
  - code: University2
    address: adress2
  - code: University3
    address: adress3
footnote:
  - code: 1
    text: "Corresponding Author"
  - code: 2
    text: "Equal contribution"

@@

Forecasting Emergency Department (ED) arrivals is a critical input to inform staffing and scheduling decisions to meet the needs of patients. An accurate ED demand forecast contributes to a better decision making process regarding the resources needed to provide services to the number and type of patients requiring hospital services. This is one of the best ways to optimize staffing utilization and related costs. Most current practice to optimise personnel scheduling follows the general approach originally presented in @vile2016time, which recommends that the following steps be taken to roster employees: (i) forecast demand; (ii) convert demand forecasts into staffing requirements; (iii) schedule shifts optimally; and (iv) assign employees to shifts. An accurate demand forecasting is crucial in ED services to depict various courses of action that can result in massive savings in terms of patient lives. Inability to match the staff with the demand results in patient overcrowding in the system which is a serious problem that causes challenging situations on patient flow [@derlet2002overcrowding]. Also, it is related with increasing length of stay [@muhammet2015forecasting], low patient satisfaction, number of patients left the hospital without seen by staff, unexpected return visits to services, increasing health care costs, inaccuracy in electronic medical record, and reported waiting times without incurring last-minute expenses, such as overtime or supplemental staffing [@rostami2020anticipating].

There exists a large literature on forecasting ED arrivals that use various methods to forecast annual[@tandberg1994time] monthly,[@chen2011long; @mai2015predicting]. daily[@rostami2020anticipating;@park2019144] and subdaily [@schweigler2009forecasting;@cheng2021forecasting] arrivals. In this paper, we only focus on forecasting hourly arrivals. In comparison to lower frequency time series forecasting such as monthly, quarterly and yearly, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series.
<!-- In this respect, forecasts of daily or monthly arrivals are likely easier but target decisions about staff allocation and the like, not the ongoing scheduling and rescheduling of how the available resources are divided among the patients in need of emergency services -->

An accurate demand forecast by hour of the day enables planners to match staff to meet anticipated patients, reconfigure units and redeploy staff. This will have many advantages for both patients, staff and the quality of provided services. Hourly forecasts are required to inform the short-term operational planning for the current and the upcoming shifts of the day. This involves the short-term decision making related to the execution of the delivery process in ED.
<!-- for various health care services such as Ambulatory, Emergency, Surgical, Inpatient, Home and Residential.  -->
The combination of an hourly forecast demand, current staff being occupied, resource availability and waiting times at ED, provide information on the state of the unscheduled care system across the service. Having this full picture enables the delivery managers to focus on the areas that require intervention to enable the most effective delivery of the service to the patients. Moreover, there are often unplanned events, such as walk-in patients, extended consultation times, during  a day or a week. Staff capacities may be adjusted based on the predicted demand fluctuations by using part-time, on-call nurses, staff overtime and voluntary absenteeism. That may also require to adjust the scheduling.

There are few studies that look at forecasting hourly arrivals in ED and other hospital services using historical time series data and/or predictors such as patient characteristics , climate factors , holidays. These studies use multiple approaches including Exponential Smoothing [REF], ARIMA families [], ARCH [REF], Vector Autoregressive [REF],  TBATS [] and ANN[]. Hourly time series generally exhibit multiple seasonal cycles of different lengths such as hourly, daily, weekly and yearly. They may also express nonstationarity and their profile may change over time. Therefore, an appropriate forecasting model should take these features into account, to accurately predict hourly demand admissions. This is currently missing in the literature and we fill this gap by examining various forecasting models capable of considering these features. All these studies only generate the estimated future arrivals as a point forecast (a single number), which does not include any uncertainty around the number of arrivals. However, reporting uncertainty is critical because the consequences of imperfect staffing are asymmetric. Therefore, probabilistic forecasts are necessary to make decisions that balance the cost associated with under and over staffing. This asymmetry arises because it is preferable to incur a small opportunity cost associated under utilised staff rather than lower service levels if staff levels are insufficient. The lack of probabilistic forecast is one of the main bottlenecks in the deployment of generated forecasts in the staffing and planning tools. In this paper, we produce and evaluate the whole forecast distribution of hourly ED arrivals, using all forecasting models, which could be used as a risk management tool for planners and decision makers. Moreover, datasets used in some of these studies are relatively small, short length (e.g. time period of 1 year), which make it challenging to report the forecast accuracy using robust approaches such as time series cross validation and such results might not be reliable. Moreover, all previous publications  covered in this paper fails reproducibility principles.

In this paper, we aim at filling these gaps and our contributions to the literature are summarised as following:

1. We develop a novel methodology to forecast short-term hourly hospital admission using the family of Generalised Additive Models that accounts for i) ..., ii) ... iii) , ... iv) (Jethro to complete)

2. We produce probabilistic forecast, in addition to the point estimation, that quantify uncertainties in future hospital admission and we evaluate its accuracy using ...;

3. We benchmark the accuracy of our model against appropriate models used when multiple seasonality is present, i.e. Prophet, TBATS, Poisson Regression, and exponential smoothing state space model (ETS);

4. We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks. The proposed approach could also be generalized to forecast hourly requirements in other services such as the number of incidents or call volumes in clinical desk services.

Our contributions are as following:

The rest of the paper is organised as following: section \ref{lit} provides a brief overview of hourly forecasting in the healthcare Section \ref{model} starts with ...

@@

There are many studies that employs models to forecast the Emergency Department needs such as admission and arrivals. The time granularity considered by these studies spans from hourly to yearly. However, given the focus of the paper, we only focus on studies on hourly ED forecasting. Table \@ref(tab:summarylit) summarize studies in hourly forecasting in the emergency department.

@@

Linear regression, ARIMA, and naive models were used by @hertzum2017forecasting to investigate whether accurate hourly accident and emergency department patient arrivals and occupancy forecasts can be generated using calendar variables. Naive model was there for the purpose of comparison. @hertzum2017forecasting study shows that patient arrivals variation is larger across the hours of the day than across the days of the week and the months of the year. In term of hour of the day, patient arrivals peaked around noon. For days of the week, Monday is the busiest day while weekends are the quietest days. July-August are the month with the highest number of patient arrivals and January and February are the months with the lowest number of arrivals. The regression and ARIMA models perform similarly for all forecast interval in modeling patient arrivals. In modeling accident and emergency department occupancy, ARIMA outperform regression models. However, after all, the models of occupancy were less accurate than those arrivals. @hertzum2017forecasting mentioned that ARIMA models are among the most accurate models for accident and emergency department visits forecasting. Another interesting point is that the accuracy of accident and emergency department forecasting models decrease with the increasing forecast interval. Lastly, the accuracy of the forecasting model may possibly be increased with additional information added to the model.

Predicting the arrivals of an ED future arrivals is also studied by @choudhury2020forecasting. ARIMA, Holt-Winters, TBATS, and neutral network methods were implemented to forecast hourly accident and emergency department arrivals. ARIMA model was selected as the best fit model. Authors claimed that ARIMA has provided high and acceptable hourly ED forecasting accuracy. @hertzum2017forecasting work was mentioned in this paper. This result is surprising given the existence of multiple seasonality in the hourly dataset.

>>>>

---
title:  Forecasting short-term hourly Emergency Departement arrivals
author:
  - name: Bahmna Rostami Tabar
    email: email1@example.com
    affiliation: University1
    footnote: 1
  - name: Jethro Browell
    email: jethro.browell@glasgow.ac.uk
    affiliation: School of Mathematics and Statistics, University of Glasgow, UK
    footnote: 2
  - name: Ivan Svetunkov
    email: i.svetunkov@lancaster.ac.uk
    affiliation: Centre for Marketing Analytics and Forecasting, Lancaster University, UK
    footnote: 2
address:
  - code: University1
    address: Cardiff business school, 3 Colum Drive, CF10 3EU, Cardiff
  - code: University2
    address: adress2
  - code: University3
    address: adress3
footnote:
  - code: 1
    text: "Corresponding Author"
  - code: 2
    text: "Equal contribution"

@@

Forecasting Emergency Department (ED) arrivals is critical for informing staffing and scheduling decisions to meet the needs of patients. An accurate ED demand forecasts contribute to a better decision making process regarding the resources allocation and staffing. This is one of the best ways to optimize resources utilisation and minimise related costs. State of the art practice to optimise personnel scheduling follows the general approach originally presented by @vile2016time, which recommends that the following steps be taken to roster employees: (i) forecast demand; (ii) convert demand forecasts into staffing requirements; (iii) schedule shifts optimally; and (iv) assign employees to shifts. An accurate demand forecasting is crucial in ED services to depict various courses of action that can result in massive savings in terms of patient lives. Inability to match the staff with the demand might result in patients overcrowding the system which is a serious problem that causes challenges for the patient flow [@derlet2002overcrowding]. Also, it is related with increasing length of stay [@muhammet2015forecasting], low patient satisfaction, unexpected return visits to services, increased health care costs, inaccuracy in electronic medical records and other [@rostami2020anticipating].

There exists a large literature on forecasting ED arrivals that use various methods to forecast annual[@tandberg1994time] monthly, [@chen2011long; @mai2015predicting] daily[@rostami2020anticipating;@park2019144] and subdaily [@schweigler2009forecasting;@cheng2021forecasting] arrivals.

<!-- In this respect, forecasts of daily or monthly arrivals are likely easier but target decisions about staff allocation and the like, not the ongoing scheduling and rescheduling of how the available resources are divided among the patients in need of emergency services -->

An accurate demand forecast by hour of the day enables planners to match staff to meet anticipated patients, reconfigure units and redeploy staff. This has many advantages for both patients, staff and the quality of provided services. Hourly forecasts are required to inform the short-term operational planning for the current and the upcoming shifts of the day. This involves the short-term decision making related to the execution of the delivery process in ED.
<!-- for various health care services such as Ambulatory, Emergency, Surgical, Inpatient, Home and Residential.  -->
The combination of an hourly forecast demand, current staff being occupied, resource availability and waiting times at ED, provide information on the state of the unscheduled care system across the service. Having this full picture enables the delivery managers to focus on the areas that require intervention to enable the most effective delivery of the service to the patients.
<!-- Moreover, there are often unplanned events, such as walk-in patients, extended consultation times, during  a day or a week. Staff capacities may be adjusted based on the predicted demand fluctuations by using part-time, on-call nurses, staff overtime and voluntary absenteeism. That may also require to adjust the scheduling. -->

However, in comparison with lower frequency time series forecasting such as monthly, quarterly and yearly, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series. Hourly time series generally exhibit multiple seasonal cycles of different lengths such as hourly, daily, weekly and yearly. They may also express nonstationarity and their profile may change over time. Therefore, an appropriate forecasting model should take these features into account, to accurately predict hourly demand admissions.
There are few studies that look at forecasting hourly arrivals in ED and other hospital services using historical time series data and/or predictors such as patient characteristics , climate factors , holidays. These studies use multiple approaches including Exponential Smoothing [REF], ARIMA families [], ARCH [REF], Vector Autoregressive model [REF],  TBATS [] and Artificial Neural Networks []. All these studies only generate the estimated future arrivals as a point forecast (a single number), which does not include any uncertainty around the number of arrivals. However, reporting uncertainty is critical because the consequences of imperfect staffing are asymmetric. This asymmetry arises because it is preferable to incur a small opportunity cost associated under utilised staff rather than lower service levels if staff levels are insufficient. The lack of probabilistic forecast is one of the main bottlenecks in the deployment of generated forecasts in the staffing and planning tools. Therefore, probabilistic forecasts are necessary to make decisions that balance the cost associated with under and over staffing. In this paper, we produce and evaluate the whole forecast distribution of hourly ED arrivals, using several forecasting models, which could be used as a risk management tool for planners and decision makers. Moreover, datasets used in some of these studies are relatively small, short length (e.g. time period of 1 year), which make it challenging to report the forecast accuracy using robust approaches such as time series cross validation and such results might not be reliable. Moreover, all previous publications  covered in this paper fails reproducibility principles.

In this paper, we aim at filling several gaps, and our contributions to the literature are summarised as follows:

1. We use the family of Generalised Additive Models to forecast short-term hourly hospital admission using  that accounts for i) ..., ii) ... iii) , ... iv) (Jethro to complete)

2. We apply advanced dynamic model based on iETS [@Svetunkov2019a] and ETSX models with a modification for multiple frequencies;

3. We produce probabilistic forecasts, in addition to the point ones, quantifying the uncertainty in future hospital admission;

4. We compare the accuracy of our models with the one from the  well established benchmarks, such as Prophet, TBATS, Poisson Regression, and exponential smoothing state space model (ETS);

5. We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks. The proposed approach could also be generalized to forecast hourly requirements in other services such as the number of incidents or call volumes in clinical desk services.

The rest of the paper is organised as following: section \ref{lit} provides a brief overview of hourly A&E arrival forecasting in healthcare. Section \ref{model} starts with ...

@@

There are many studies that employ models to forecast the admission and arrivals in Emergency Department. The time granularity considered by these studies spans from hourly to yearly. However, given the focus of the paper, we only discuss the hourly ED forecasting. Table \@ref(tab:summarylit) summarize the relevant papers.

@@

**Gijo paper is missing in the table above!**

Linear regression, ARIMA, and Naïve were used by @hertzum2017forecasting to investigate whether accurate hourly accident and emergency department patient arrivals and occupancy forecasts can be generated using calendar variables. Naïve method was there for the purpose of comparison. @hertzum2017forecasting study shows that patient arrivals variation is larger across the hours of the day than across the days of the week and the months of the year. In terms of hour of the day, patient arrivals peaked around noon. For days of the week, Monday is the busiest day while weekends are the quietest ones. July-August are the months with the highest numbers of patient arrivals, while January and February are the months with the lowest numbers. According to @hertzum2017forecasting, the regression and ARIMA models performed similarly in modelling patient arrivals, while ARIMA outperformed regression models in modelling accident and emergency department occupancy. However, after all, the models of occupancy were less accurate than those of arrivals. @hertzum2017forecasting mentioned that ARIMA was among the most accurate models in terms of forecasting accident and emergency department visits.
<!-- Another interesting point is that the accuracy of accident and emergency department forecasting models decrease with the increasing forecast interval. Lastly, the accuracy of the forecasting model may possibly be increased with additional information added to the model. -->

@choudhury2020forecasting used ARIMA, Holt-Winters, TBATS, and neutral network methods to forecast hourly accident and emergency department arrivals. ARIMA model was selected as the best fit model. Authors claimed that ARIMA has provided high and acceptable hourly ED forecasting accuracy, even outperforming TBATS.
<!-- @hertzum2017forecasting work was mentioned in this paper. This result is surprising given the existence of multiple seasonality in the hourly dataset. -->
